---
title: "LAB 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=FALSE, warning=FALSE, results='hide'}
#packages needed for the lab
library(kknn)
library(readxl)
```

Exercise 1

```{r, warning=FALSE, results='hide', echo=FALSE}

data = read.csv("optdigits.csv", header = F)
n = dim(data)[1]
set.seed(12345)


id=sample(1:n, floor(n*0.5))
train=data[id,]

id1=setdiff(1:n, id)
set.seed(12345)

id2=sample(id1, floor(n*0.25))
valid=data[id2,]

id3=setdiff(id1,id2)
test=data[id3,]


#Exercise 2


k_test <- kknn(as.factor(V65)~., train = train, test = test, k = 30, kernel = "rectangular")
k_train <- kknn(as.factor(V65)~., train = train, test = train, k = 30, kernel = "rectangular")
k_valid <- kknn(as.factor(V65)~., train = train, test = valid, k = 30, kernel = "rectangular")

test_table = table(k_test$fitted.values, as.factor(test$V65))
train_table = table(k_train$fitted.values, train$V65)
valid_table = table(k_valid$fitted.values, valid$V65)

missclass = function(X, X1) {
  n = length(X)
  return (1 - sum(diag(table(X1, X)))/n)
}

missclass_test = missclass(k_test$fitted.values, as.factor(test$V65))
missclass_train = missclass(train$V65, k_train$fitted.values)
missclass_valid = missclass(k_valid$fitted.values, valid$V65)
```

Confusion matrices

```{r, echo=T}
test_table
train_table 
```

Missclass Values
```{r, echo=T}
missclass_test
missclass_train


```

#Comment on the quality of predictions for different digits and on the overall prediction quality.

#Some numbers have worse predictions than others, for example number 4 was predicted to be a 7 a couple of times. 
#This is most likely due to slopy writing that makes different numbers look more similar to others. 
#The overall prediction quality is good.
Exercise 3

```{r, echo=F}

prob_eight <- k_train$prob[, 9]

ordered_high <- order(prob_eight, decreasing = T)
ordered_high <- ordered_high[1:2]

ordered_low <- order(prob_eight, decreasing = F)
ordered_index <- which(prob_eight > 0)

ordered_low_eights <- c()
index = 1
while(length(ordered_low_eights) < 3) {
  value <-train[ordered_low[index], 65]
  if (value == 8) {
    ordered_low_eights <- append(ordered_low_eights, ordered_low[index])
  }
  index <- index + 1
}

print(ordered_low_eights)

  
  ordered_matrix <- c(ordered_low_eights, ordered_high)

for (i in ordered_matrix){
  my_heatmap <- matrix(as.numeric(train[i,1:64]), nrow=8,ncol=8, byrow = T)
  heatmap(my_heatmap, Colv = NA, Rowv = NA, main =train[i, 65])
}

```


#comment on whether these cases seem to be hard or easy to recognize visually. 

#The eights that were easy to classify are also very easy to see as eights. 
#The eights that were most difficult to spot are very difficult to see without knowing that they're eights beforehand. 


Exercise 4

```{r, echo=F}
train_miss_error <- numeric(30)
val_miss_error <- numeric(30)


for (i in 1:30){
  k_valid <- kknn(as.factor(V65)~., train = train, test = valid, k = i, kernel = "rectangular")
  k_train <- kknn(as.factor(V65)~., train = train, test = train, k = i, kernel = "rectangular")
  
  train_miss_error[i] <- missclass(k_train$fitted.values, train$V65)
  val_miss_error[i] <- missclass(k_valid$fitted.values, valid$V65)
  
}

plot = plot(c(1:30), train_miss_error, ylab = "train_miss_error", xlab = "k", col="pink")
points(c(1:30), val_miss_error, col="green")

k_test <- kknn(as.factor(V65)~., train = train, test = test, k = 4, kernel = "rectangular")
k_train <- kknn(as.factor(V65)~., train = train, test = train, k = 4, kernel = "rectangular")
k_valid <- kknn(as.factor(V65)~., train = train, test = valid, k = 4, kernel = "rectangular")

missclass_test = missclass(k_test$fitted.values, as.factor(test$V65))
missclass_train = missclass(train$V65, k_train$fitted.values)
missclass_valid = missclass(k_valid$fitted.values, valid$V65)
```



Missclass Values
```{r, echo=T}
missclass_test
missclass_train
missclass_valid


```

#How does the model complexity change when K increases and how does it affect the training and validation errors?

#The model complexity remains the same as k increases, since it's not learning more parameters or features as k increases.
#One can argue that the overall complexity is increased since there are more data points added, but this does not mean that the complexity of the model
#is higher. 
#The best K values are the ones with the lowest missclass error, in this case k=3 and k=4 are the best values. 

#We can see that train gives the lowest missclass error because it is the date we made the model from.
#Both test and validation is slightly higher which is to be expected.
#With the right k value computed with the validation data we get a test error which is just above 2.5% 
#Which results in a pretty high quality model.


Exercise 5

```{r, echo=F}
cross.entropy <- function(p, phat){
  x <- 0
  for (i in 1:length(p)){
    x <- x + (p[i] * log(phat[i]))
  }
  return(-x)
}

train_miss_error <- as.vector(matrix(0,ncol = 30))
val_miss_error <- as.vector(matrix(0,ncol = 30))

entropy_error <- as.vector(matrix(0,ncol = 30))

for (i in 1:30){
  k_valid <- kknn(as.factor(V65)~., train = train, test = valid, k = i, kernel = "rectangular")
  k_train <- kknn(as.factor(V65)~., train = train, test = train, k = i, kernel = "rectangular")
  
  
  for (j in 0:9){
    cross_val <- valid$V65 == j
    cross_train <- train$V65 == j
    
    bool_val <- (which(cross_val, useNames = T))
    prob_val <- k_valid$prob[cross_val, as.character(j)] + 1e-15
    
    bool_train <- (which(cross_train, useNames = T))
    prob_train <- k_train$prob[cross_train, as.character(j)] + 1e-15
    
    #val_miss_error[i] <- val_miss_error[i] + sum(-(bool_val*log(prob_val)))
    #train_miss_error[i] <- train_miss_error[i] + sum(-(bool_train*log(prob_train)))
    
    val_miss_error[i] <- cross.entropy(bool_val, prob_val)
    train_miss_error[i] <- cross.entropy(bool_train, prob_train)
    
    
    
    entropy_error[i] <- abs(val_miss_error[i]-train_miss_error[i])
  }
  
}

plot(c(1:30), entropy_error, ylab = "cross_entropy_error", xlab = "k", col="blue")
which.min(entropy_error)

```


#What is the optimal ð¾ð¾ value here? Assuming that response has multinomial distribution, why might the cross-entropy be a more suitable choice of the error function than the missclassification error for this problem?

#This model might be more suitable because of lower complexity levels than missclass error.
#The optimal K value here is K = 5. 

Assigment 2


```{r, echo=F}
library(caret)
parkin = read.csv("parkinsons.csv", header = T)
parkin_corr <- data.frame(parkin[c(5,7:22)]) #Remove unused voice characteristics
parkin_scaled <- as.data.frame(scale(parkin_corr))

n = dim(parkin_corr)[1]
set.seed(12345)

id=sample(1:n, floor(n*0.6))

train_pre=parkin_corr[id,]
test_pre=parkin_corr[-id,]

data_scaler<-preProcess(train_pre)
train<-predict(data_scaler,train_pre)
test<-predict(data_scaler,test_pre)
```
Excercise 2

```{r, echo=T}
fit = lm(motor_UPDRS ~ .-1, data = train)
sum = summary(fit)
MSE_train<-mean(sum$residuals^2)
pred_test<-predict(fit,test)
MSE_test<-mean((test$motor_UPDRS-pred_test)^2)
print(MSE_train)
print(MSE_test)
print(sum)


```

#The variables p-values that are higher than 0.05 does not contribute at all (insignificant P-values).
#The values that contribute significantly are the ones with low p-values.
#The variables that will contribute significantly to the model are the ones with the highest amount of stars in the table,
#which represents the lowest p-values. 

#MSE_train = 0.8785431
#MSE_test = 0.9354477

Exercise3

```{r, echo=F}
log_likelihood <- function(train, Y, theta, sigma){

  n <- dim(train)[1] 
  train_theta = train%*%theta

  sum1 <- n*log(sigma^2)/2
  sum2 <- n*log(2*pi)/2
  sum3 <- sum((train_theta-Y)^2) 
  sum4 <- sum3/(2*sigma^2)
  
  return (-sum1-sum2-sum4)
}


ridge <- function(train, theta, lambda, Y){
  n<-dim(train)[2]
  sigma <- theta[n+1]
  theta <-as.matrix(theta[1:n])
  log_like <- log_likelihood(theta=theta,Y=Y,sigma=sigma,train=train)
  ridge <- -log_like + lambda*sum(theta^2)
  return(ridge)
  
}                                     

ridgeOpt <- function(lambda, train, Y){
  
  train <- as.matrix(train)
  N = dim(train)[2] 
  init_theta = integer(N)
  init_sigma = 1
  
  opt <- optim(par = c(init_theta,init_sigma), fn = ridge, lambda = lambda, train = train, Y = Y, method = "BFGS")
  return(opt)
} 

dF <- function(X, lambda){
  #From the course formula
  X <- as.matrix(X)
  Xt <- t(X)
  n <- dim(X)[2]
  I <- diag(n)
  P <- X%*%solve((Xt%*%X + (lambda*I)))%*%Xt
  return(sum(diag(P)))
} 
```

```{r, echo=F}
  AIC = function(train,Y,theta, sigma, lambda){
    log_like = log_likelihood(train = train, Y=Y, theta = theta, sigma = sigma)
    N = dim(train)[1] # No of data points
    df = dF(train,lambda)
    aic = (-2*log_like/N) + (2*df/N) #(-2*Log-likelihood/N) + 2*(df/N)
    return(aic)
  }
```

Exercise 4

```{r, echo=F}
xtrain<-as.matrix(train[2:17])
ytrain<-as.matrix(train[1])
xtest=as.matrix(test[2:17])
ytest<-as.matrix(test[1])


for (lambda in c(1, 100, 1000)){
  opt = ridgeOpt(lambda, xtrain, ytrain)
  theta <- as.matrix(opt$par[1:16]) 
  sigma<- opt$par[17]
  MSE_train = mean((xtrain%*%theta - ytrain)^2)
  MSE_test = mean((xtest%*%theta - ytest)^2)
  aic = AIC(train=xtrain,Y= ytrain,theta = theta,sigma= sigma, lambda= lambda)
  print(paste("Lambda:",lambda))
  print(paste("TrainMSE:",MSE_train))
  print(paste("TestMSE:",MSE_test))
  print(paste("AIC:", aic)) #Unecessary


}

```

#Which penalty parameter is most appropriate among the selected ones? Compute and compare the degrees of freedom of these models and make appropriate conclusions.

#We have read that to find the best aic value among mutiple aic values is the lowest. In our example it seems like lambda = 1 and lambda = 100 gives similar aic values which is hard to draw conclusions from. Lambda = 1 gives the lowest AIC value.
#Since the the Lambda = 100 has the lowest test_MSE, this is the best value. 
#The degrees of freedom reflects the complexity of the model, so the larger lambda, the smaller degrees of freedom we would get. We get more complex models, basically penalizing the complexity. 


Assignment 3.

```{r, echo=F}
prime = read.csv("pima-indians-diabetes.csv", header = F)

set.seed(12345)

```


```{r, echo=F}
coloor <- function(x){
  if (x==1){
    c = "red"
  } else{
    c="green"
  } 
  return(c)
}  

coloors = sapply(prime$V9, coloor)
plot( prime$V2, prime$V8, xlab = "Plasma", ylab = "Age", main = "doabets", col = coloors)
```

seems to be different, no obvious correlation between the variables.
Making a linear regression would not provide a meaningful line. 


Exercise 2

```{r, echo=F}
glm.fits = glm(V9~ V2 + V8, prime, family = "binomial" )
prob=predict(glm.fits, type="response")
pred=ifelse(prob>0.5, 'red','green')

table = table(pred, prime$V9)

miss <- missclass(pred, prime$V9)

plot(prime$V2, prime$V8,col=pred, xlab = "Plasma glucose levels", ylab = "Age", main = paste("Missclass error", toString(miss), sep=" = ") ) 

```


It seems to be an acceptable classification with about 1/4 error but not perfect 
There seems to be a division when you reach above 150 in plasma-glucose levels. But this threshold seems to lower as you age. An older person is more likely to have diabetes even with lower plasma-glucose levels. 

Exercise 3

```{r, echo=F}

r=.5
glm.fits = glm(V9~ V2 + V8, prime, family = "binomial" )
cf = glm.fits$coefficients
prob=predict(glm.fits, type="response")
pred=ifelse(prob>0.5, 'red','green')

w9 = cf[1]
w2 = cf[2]
w8 = cf[3]

x8 = c(seq(0,100,0.1))
x2 = (log(-r/(r-1)) - w9 - w8*x8)/w2

plot(prime$V2, prime$V8,col=pred, ylab = "Age", xlab= "Plasma", main = paste("Missclass Error", toString(miss), sep=" = ")) 
lines(x2,x8,col="blue")
```

It catches the distribution very well


Exercise 4

```{r, echo=F}
for(r in c(.2,.5, .8)) {
  pred=ifelse(prob>r, 'red','green')
  table(pred, prime$V9)
  
  miss <- missclass(pred, prime$V9)
  
  x2 = (log(-r/(r-1)) - w9 - w8*x8)/w2
  
  plot(prime$V2, prime$V8,col=pred, ylab = "Age", xlab= "Plasma", main = paste("Missclass_Error = ", toString(miss), "\n r = ", r, sep= "")) 
  lines(x2,x8,col="blue")
}
```

With r=.2 and r=.8, the missclassification error increased. r=.2 was clearly worse (37% error). 



Exercise 5


```{r, echo=F}
expanded <- prime

expanded$z1 <- expanded$V2 ** 4
expanded$z2 <- expanded$V2 ** 3 * expanded$V8
expanded$z3 <- expanded$V2 ** 2 * expanded$V8 ** 2
expanded$z4 <- expanded$V2 * expanded$V8 ** 3
expanded$z5 <- expanded$V8 ** 4

glm.fits = glm(V9~ V2 + V8 + z1 + z2 + z3 + z4 + z5, expanded, family = "binomial" )


for(r in c(.2,.5, .8)) {
  prob=predict(glm.fits, type="response")
  pred=ifelse(prob>r, 'red','green')
  table(pred, expanded$V9)
  
  miss <- missclass(pred, prime$V9)
  
  x2 = (log(-r/(r-1)) - w9 - w8*x8)/w2
  
  plot(expanded$V2, expanded$V8,col=pred, ylab = "Age", xlab= "Plasma", main = paste("Missclass_Error = ", toString(miss), "\n r = ", r, sep= "")) 
}
```


This model has a lower missclass error, so the quality of this model is slightly better than the previous one. 
The linearity of the model is however gone, since we've added non-linear variables. This causes the plotted colors to look more like a "slice"



