---
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE}


library(kernlab)
set.seed(1234567890)

data(spam)
foo <- sample(nrow(spam))
spam <- spam[foo,]
spam[,-58]<-scale(spam[,-58])
tr <- spam[1:3000, ]
va <- spam[3001:3800, ]
trva <- spam[1:3800, ]
te <- spam[3801:4601, ] 

by <- 0.3
err_va <- NULL
for(i in seq(by,5,by)){
  filter <- ksvm(type~.,data=tr,kernel="rbfdot",kpar=list(sigma=0.05),C=i,scaled=FALSE)
  mailtype <- predict(filter,va[,-58])
  t <- table(mailtype,va[,58])
  err_va <-c(err_va,(t[1,2]+t[2,1])/sum(t))
}

filter0 <- ksvm(type~.,data=tr,kernel="rbfdot",kpar=list(sigma=0.05),C=which.min(err_va)*by,scaled=FALSE)
mailtype <- predict(filter0,va[,-58])
t <- table(mailtype,va[,58])
err0 <- (t[1,2]+t[2,1])/sum(t)
err0

filter1 <- ksvm(type~.,data=tr,kernel="rbfdot",kpar=list(sigma=0.05),C=which.min(err_va)*by,scaled=FALSE)
mailtype <- predict(filter1,te[,-58])
t <- table(mailtype,te[,58])
err1 <- (t[1,2]+t[2,1])/sum(t)
err1

filter2 <- ksvm(type~.,data=trva,kernel="rbfdot",kpar=list(sigma=0.05),C=which.min(err_va)*by,scaled=FALSE)
mailtype <- predict(filter2,te[,-58])
t <- table(mailtype,te[,58])
err2 <- (t[1,2]+t[2,1])/sum(t)
err2

filter3 <- ksvm(type~.,data=spam,kernel="rbfdot",kpar=list(sigma=0.05),C=which.min(err_va)*by,scaled=FALSE)
mailtype <- predict(filter3,te[,-58])
t <- table(mailtype,te[,58])
err3 <- (t[1,2]+t[2,1])/sum(t)
err3

```

# Questions

# 1. Which filter do we return to the user ? filter0, filter1, filter2 or filter3? Why?

The filter we return is filter1. Because the SVM model is trained on trianing data (tr) and validated on validation data (va). Then we should use the optimal err_va to make prediction on the test data. Using ksvm on training data and not any other. In general you do your training on your training set, evalutate its performance on the validation set, and then use the best model to predict on the test dataset. 

# 2. What is the estimate of the generalization error of the filter returned to the user? err0, err1, err2 or err3? Why?

it is err1 = 0.08489388. Filter0 gives low error estimate due to the fact that it is filtered on validation, the same way we choose the optimal err_va. Filter3 gives the lowest error estimate because of the the model is trained on the same data it is predicted on. As for filter1 and filter2, they give similar error estimates. Filter2 error rate is slightly lower due to the fact that it is trained on a larger data set (training and validation), where filter1 is only trained on training data.

# 3. Implementation of SVM predictions.


```{r}
# Get the indices of the support vectors in the SVM classifier
sv <- alphaindex(filter3)[[1]]

# Get the coefficients of the support vectors in the SVM classifier
co <- coef(filter3)[[1]]

# Get the bias term of the SVM classifier and negate it
inte <- -b(filter3)

# Create an RBF kernel with a standard deviation of 0.05
rbfkernel <- rbfdot(sigma = 0.05)

# Initialize an empty vector to store the predicted values
k <- c()

# Loop over the first 10 rows of the spam dataset
for(i in 1:10) { 
  
  # Initialize a variable to store the predicted value for the current row
  k2 <- 0
  
  # Loop over each support vector
  for(j in 1:length(sv)) {
    
    # Apply the RBF kernel to the jth support vector and the ith row of the spam dataset
    f <- rbfkernel(unlist(spam[sv[j], -58]), unlist(spam[i, -58]))
    
    # Update the predicted value for the current row by adding the product of the jth coefficient and the value of the kernel
    k2 <- k2 + co[j] * f[1]
  }
  
  # Append the predicted value for the current row to the k vector
  k <- c(k, k2 + inte)
}

# Print the k vector
k

# Use the predict function to make predictions on the first 10 rows of the spam dataset using the trained SVM classifier
prediction = predict(filter3, spam[1:10, -58], type = "decision")

plot(k, col = "red")
lines(prediction)

```
The red dotted points is the predicted values that generated from the linear combination for filter3. The lines is the predictions from the predict function. 



